Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-05-30 doc/thesis/methods_results.tex
A bit more writing

@@

The accuracy of grid search estimates largely paralleled that of the \gls{kSVR}
classifiers. \Cref{fig:gridest} shows point estimates and 95\% highest density
intervals for each of the \gls{BA} parameters, for one replicate experiment
with 500-tip trees. Plots showing the point estimates for all replicates can be
found in \cref{fig:gridptalpha,fig:gridptI,fig:gridptm,fig:gridptN}. For all
parameters except $m$, the error of point estimates was negatively correlated
with the number of sampled tips in the tree (for
\gls{alpha}, \gls{I}, and \gls{N} respectively: Spearman's $\rho$ = 
    \ensuremath{-0.22},
    \ensuremath{-0.51},
    \ensuremath{-0.16};
$p$-values
    $4\!\times\!10^{-4}$,
    ${<}10^{-5}$,
    $0.01$).
The highest density intervals obtained for all parameters were extremely wide,
occupying $>$75\% of the grid in all cases (\cref{fig:gridest}).

The \gls{alpha} parameter was the most accurately estimated, with point
estimates having an average deviation of 
    0.14
from the true value, on a grid from 0 to 2. The error of point estimates varied
significantly between true values of \gls{alpha}
    (one-way \gls{ANOVA}, $p$ ${<}10^{-5}$). In
particular, errors were lower for the values \gls{alpha} = 1.0 and 1.25 than
for the other values
    (average errors 
    0.03
    for \gls{alpha} = 1.0 or 1.5 vs.
    0.17
    for \gls{alpha} $\neq$ 1.0 or 1.5),
and this difference was significant
    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
     \cref{fig:gridptalpha}).
These two values exhibited different qualitative behaviour than the other
values in terms of the distribution of kernel scores along the grid
(\cref{fig:gridalpha}). In particular, there was a pronounced peak in scores
around the true value, in contrast to the other values where the scores were
flat around the true value. The effect was most obvious for the value
\gls{alpha} = 1.25.

The average absolute error of the point estimates for \gls{I} was 
    310 individuals,
on a grid of 500 to 5000, and these errors differed between true values of
\gls{I}
    (one-way \gls{ANOVA}, $p =0.001$).
The errors for $2000 \leq I \leq 3000$ were higher than those for the other
values
    (average errors
     430
     for $2000 \leq I \leq 3000$ vs.
     250
     for $I < 2000$ or $I > 3000$),
and this difference was significant
    (Wilcoxon rank-sum test, $p =6\!\times\!10^{-4}$,
     \cref{fig:gridptI}).
Kernel score distributions for all test values exhibited a similar rounded
shape (\cref{fig:gridI}).

The average error for \gls{m} was
    1.31 edges per vertex,
on a grid from 1 to 6. The error varied significantly between the true values
of \gls{m} 
    (one-way \gls{ANOVA}, $p {<}10^{-5}$).
Errors for the value \gls{m} = 1 were lower than the other values
    (average errors
    0.1
    for $m = 1$ vs.
    1.55
    for $m > 1$),
and this difference was significant
    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
     \cref{fig:gridptm}).
The value $m = 1$ causes the network to take on a distinct shape relative to
higher \gls{m} values, namely a tree (\ie there are no cycles, see
\cref{subsec:treeshape}). The kernel score distribution had a peak at $m = 1$
when this was the true value, and a valley at $m = 1$ when the true value of
$m$ was greater that 1 (\cref{fig:gridm}).

The average error for \gls{N} was 
    2419 individuals,
on a grid from 1000 to 15000, and was varied significantly with the true value
of \gls{N}
    (one-way \gls{ANOVA}, $p {<}10^{-5}$).
The errors were lower for $N \leq 3000$
    (average errors
     740
     for $N \leq 3000$ vs.
     2979
     for $N > 3000$),
and this difference was significant
    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
     \cref{fig:gridptN}).
The kernel score distribution had a peak at $N = 1000$ when this was the true
value, and a valley there otherwise (\cref{fig:gridN}). Except for this valley,
the distributions were flat for $N > 3000$.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{gridsearch-example}
  \caption[Grid search estimates of \gls{BA} model parameters.]{Point estimates
      and 95\% highest density intervals for each \gls{BA} model parameter,
      obtained using grid search. Networks and transmission trees were
      simulated over a grid of values for each parameter while holding the
      others fixed. For a subset of the grid values ($x$-axis), test networks
      and trees were created and compared to each tree on the grid using the
      tree kernel. The kernel scores along the grid were normalized to resemble
      a probability distribution, from which the mode and highest density
      interval were calculated. Shown values correspond to one replicate
      experiment, with trees of size 500.
  } 
  \label{fig:gridest}
\end{figure}

\subsubsection*{Joint parameter estimates with kernel-assisted ABC}

\Cref{fig:abcptm2} shows {\color{red}\sout{\gls{MAP}}}
{\color{blue}\uline{posterior mean}} point estimates of the BA model parameters
obtained with kernel-assisted ABC on simulated data. The estimates shown
correspond only to the simulations where the $m$ parameter was set to 2,
however the results for $m = 3$ and $m = 4$ were similar
(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% HPD intervals are
given in \cref{tab:abchpd}.

The accuracy of the parameter estimates obtained with kernel-assisted ABC
paralleled the results from the \gls{kSVR} classifier. Of the four parameters,
$\alpha$ was the most accurately estimated, with point estimates having a
median [IQR] absolute error of 
    0.11 
    [0.03 - 
    0.25].
The errors when the true value of $\alpha$ was zero were significantly greater
than those for the other values 
    (Wilcoxon rank-sum test, $p$ = $0$).
Errors in estimating $\alpha$ 
%also varied with the true value of $m$ just at
%the threshold of statistical significance
%    (one-way ANOVA, $p 
%    pp(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"], eq=TRUE)$),
{\color{blue}\uline{did not vary across the true values of $m$ or $I$ (both
one-way ANOVA).}}

Estimates for $I$ were relatively accurate, with point estimate errors of
    492 
    [294 - 
    782] individuals.
These errors were significantly higher when the true value of $\alpha$ was
at least 1
    (Wilcoxon rank-sum test, $p$ = $0$)
and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
did not affect the estimates of $I$ (one-way ANOVA).

The $m$ parameter was estimated correctly in
    37 \%
of simulations {\color{red}\sout{barely better than random guessing}}.
{\color{blue}\uline{However this was due to the estimated value being $m = 3$
in the large majority of simulations (62 out of
72), with one-third of the simulations having the true value $m =
3$.}} The true values of the other parameters did not significantly affect the
estimates of $m$ (both one-way ANOVA).

Finally, the total number of nodes $N$ was consistently over-estimated by about
a factor of two (error 4153
[3660 - 4489] individuals). No parameters influenced the accuracy of the $N$
estimates (all one-way ANOVA).

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{abc-point-estimate-m2}
  \vspace{6pt}
  \caption[
    Posterior mean point estimates for \gls{BA} model parameters obtained by
    running \software{netabc} on simulated data, for simulations with \gls{m} =
    2.
  ]{
    Posterior mean point estimates for \gls{BA} model parameters obtained by
    running \software{netabc} on simulated data, for simulations with \gls{m} =
    2. Dashed lines indicate true values. (A) Estimates of \gls{alpha} and
    \gls{I} which were varied in these simulations against known values. (B)
    Estimates of \gls{m} and \gls{N} which were held fixed in these simulations
    at the values \gls{m} = 2 and \gls{N} = 5000.
  }
  \label{fig:abcptm2}
\end{figure}

\begin{table*}[ht]
  \centering
  \input{\tablepath/abc-hpd.tex}
  \caption[
      Average posterior mean point estimates and 95\% highest posterior density
      (HPD) interval widths for BA model parameter estimates obtained with
      kernel-assisted ABC.
  ]{
      Average posterior mean point estimates and 95\% highest posterior density
      (HPD) interval widths for BA model parameter estimates obtained with
      kernel-assisted ABC. Three transmission trees were simulated under each
      combination of the listed parameter values, and the parameters were
      estimated with kernel-assisted ABC without training.
  }
  \label{tab:abchpd}
\end{table*}

The dispersion of the ABC approximation to the posterior also varied between
the parameters, with narrower HPD intervals for the parameters with more
accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows
the distributions for one simulation. Equivalent plots for one replicate
simulation with each studied parameter combination can be found in
\cref{fig:0.0-1000-2-5000-0,fig:0.5-1000-2-5000-0,fig:1.0-1000-2-5000-0,fig:1.5-1000-2-5000-0,fig:0.0-2000-2-5000-0,fig:0.5-2000-2-5000-0,fig:1.0-2000-2-5000-0,fig:1.5-2000-2-5000-0,fig:0.0-1000-3-5000-0,fig:0.5-1000-3-5000-0,fig:1.0-1000-3-5000-0,fig:1.5-1000-3-5000-0,fig:0.0-2000-3-5000-0,fig:0.5-2000-3-5000-0,fig:1.0-2000-3-5000-0,fig:1.5-2000-3-5000-0,fig:0.0-1000-4-5000-0,fig:0.5-1000-4-5000-0,fig:1.0-1000-4-5000-0,fig:1.5-1000-4-5000-0,fig:0.0-2000-4-5000-0,fig:0.5-2000-4-5000-0,fig:1.0-2000-4-5000-0,fig:1.5-2000-4-5000-0}. HPD intervals around $\alpha$ and $I$ were often narrow
relative to the region of nonzero prior density, whereas the intervals for $m$
and $N$ were more widely dispersed.

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{abc-posterior-example}
  \vspace{6pt}
  \caption{
    Marginal posterior distributions of BA model parameters estimated
    with kernel-assisted ABC for a single simulated transmission tree. Dotted
    lines and shaded polygon indicate true values.
  }
  \label{fig:abcex}
\end{figure}

To test the effect of model misspecification, we simulated one network where
the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The posterior mean
[95\% HPD] estimates for each parameter were: 
$\alpha$, 
  1.09 
  [0.6 -
   1.14];
$I$,
  1143 
  [666 -
   4455];
$m$,
  3 
  [2 -
   5];
$N$,
  12635 
  [4282-
   14977].
The approximate posterior distributions for this simulation are shown in
\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
transmission tree in a peer-driven fashion, where the probability to sample a
node was twice as high if one of its peers had already been sampled. The
parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
$I$ = 2000. The estimated values were
$\alpha$, 
  0.23 
  [0 -
   0.63];
$I$,
  2383 
  [1419 -
   3811];
$m$,
  4 
  [2 -
   5];
$N$,
  13517 
  [2836 -
   14720].
The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
of these results were in line with estimates obtained on other simulated
datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
$\alpha$ was somewhat lower than typical.

%<<abc_glm, include=FALSE>>=
%    source("global.R")
%    options(scipen=-1, digits=2)
%@
%
%<<alpha_glm, include=FALSE>>=
%    # alpha_error is influenced by alpha and m but not I
%    stopifnot(alpha.glm[Parameter == "alpha", min(p) < 0.05])
%    stopifnot(alpha.glm[Parameter == "m", min(p) < 0.05])
%    stopifnot(alpha.glm[Parameter == "I", min(p) > 0.05])
%
%    # alpha_error is correlated with true_alpha
%    alpha.test <- d[,cor.test(alpha_error, true_alpha, method="spearman")]
%    stopifnot(alpha.test$p.value < 0.05)
%
%    # alpha_error is not correlated with true_m
%    alpha.m.cor <- d[,cor.test(alpha_error, true_m)]
%    stopifnot(alpha.m.cor$p.value < 0.05)
%@
%
%We used \software{netabc} to estimate the parameters of the \gls{BA} model on
%simulated trees where the true parameter values were known. Point estimates for
%each parameter are shown in \cref{fig:abcptm2} for the simulations with \gls{m} =
%2. The results for the other values of \gls{m} were similar
%(\cref{fig:abcptm3,fig:abcptm4}). The median [IQR] absolute error of estimates
%of \gls{alpha} across all simulations was
%    d[,median(alpha_error)] 
%    [d[,quantile(alpha_error, 0.25)]-d[,quantile(alpha_error, 0.75)]].
%\Gls{GLM} analysis indicated that the true values of both \gls{alpha} and
%\gls{m} had significant effects on the error in estimated \gls{alpha}
%    ($p$ values $pp(alpha.glm[Parameter == "alpha", min(p)], eq=FALSE)$
%     $pp(alpha.glm[Parameter == "m", min(p)], eq=FALSE)$,
%     \cref{tab:glmalpha}),
%but the true value of \gls{I} did not. There was a significant negative
%correlation between the true value of \gls{alpha} and the error
%    (Spearman's $\rho$ = round(alpha.test$estimate, 2),
%     $p pp(alpha.test$p.value, eq=TRUE)$).
%There was a negative correlation between \gls{m} and the error in \gls{alpha}
%    (Spearman's $\rho$ = round(alpha.m.cor$estimate, 2),
%    $p pp(alpha.m.cor$p.value, eq=FALSE)$).
%
%<<I_glm, include=FALSE>>=
%    # I_error is influenced by alpha and I but not by m
%    stopifnot(I.glm[Parameter == "alpha", min(p) < 0.05])
%    stopifnot(I.glm[Parameter == "m", min(p) > 0.05])
%    stopifnot(I.glm[Parameter == "I", min(p) < 0.05])
%
%    # correlation between I_error and true_alpha
%    I.alpha.cor <- d[,cor.test(I_error, true_alpha, method="spearman")]
%    stopifnot(I.alpha.cor$p.value < 0.05)
%
%    # correlation between I_error and true_I
%    I.I.cor <- d[,cor.test(I_error, true_I, method="spearman")]
%    stopifnot(I.alpha.cor$p.value < 0.05)
%@
%
%The mean error in the estimated value of \gls{I} was
%    d[,round(median(I_error))] 
%    [d[,round(quantile(I_error, 0.25))]-d[,round(quantile(I_error, 0.75))]],
%an over-estimate of roughly a factor of 1.5
%(\cref{fig:abcptm2,fig:abcptm1,fig:abcptm3,fig:abcptm4}). \Gls{GLM} analysis
%indicated a relationship between the error in estimated \gls{I}, and the true
%values of \gls{alpha} and \gls{I}
%    ($p$ values $pp(I.glm[Parameter == "alpha", min(p)], eq=FALSE)$
%     $pp(alpha.glm[Parameter == "I", min(p)], eq=FALSE)$,
%     \cref{tab:glmalpha}),
%but not the true value of \gls{m}. There was a significant correlation between 
%the true value of \gls{alpha} and the error in estimated \gls{I}
%    (Spearman's $\rho$ = I.alpha.cor$estimate,
%     $p pp(I.alpha.cor$p.value, eq=TRUE)$),
%and between the true value of \gls{I} and the error in estimated \gls{I}
%    (Spearman's $\rho$ = I.I.cor$estimate,
%     $p pp(I.I.cor$p.value, eq=TRUE)$).
%
%<<m_glm, include=FALSE>>=
%    # effects of parameters on m_error
%    stopifnot(m.glm[Parameter == "m", min(p)] < 0.05)
%    stopifnot(m.glm[Parameter == "alpha", min(p)] > 0.05)
%    stopifnot(m.glm[Parameter == "I", min(p)] > 0.05)
%
%    # m was only estimated correctly in just over 20\% of simulations with m > 2
%    m.tbl <- prop.table(table(d[true_m > 1,m_error]))
%    stopifnot(m.tbl[1] > 0.2 & m.tbl[1] < 0.3)
%
%    # with m = 1, over 95% of simulations are right
%    m1.est <- d[true_m == 1, m_error]
%    stopifnot(sum(m1.est == 0) / length(m1.est) > 0.95)
%    stopifnot(sum(m1.est == 0) / length(m1.est) < 1)
%@
%
%\gls{GLM} analysis showed an effect of the true value of \gls{m} on the error
%in the estimated \gls{m}
%    ($p pp(m.glm[Parameter == "m", min(p)])$,
%     \cref{tab:glmm}).
%When the true value of \gls{m} was 1, the correct value was recovered by
%\software{netabc} in virtually every case
%    (sum(m1.est == 0) out of length(m1.est) simulations).
%However, when the true value of \gls{m} was 2 or higher, the correct value was
%recovered in only 
%    as.integer(m.tbl[1] * 100)\%
%of simulations, little better than random guessing. The \gls{GLM} analysis did
%not indicate any effects of the true parameter values on the error in estimated
%\gls{m} (\cref{tab:glmm}).
%
%<<N_glm, include=FALSE>>=
%    # alpha_error is influenced by alpha and m but not I
%    stopifnot(N.glm[Parameter == "alpha", min(p) < 0.05])
%    stopifnot(N.glm[Parameter == "I", min(p) < 0.05])
%    stopifnot(N.glm[Parameter == "m", min(p) > 0.05])
%
%    N.alpha.test <- d[,wilcox.test(.SD[true_alpha == 1.5, N_error], 
%                                   .SD[true_alpha < 1.5, N_error])]
%    stopifnot(N.alpha.test$p.value < 0.05)
%    N.I.test <- d[,wilcox.test(.SD[true_I == 3000, N_error], 
%                               .SD[true_I < 3000, N_error])]
%    stopifnot(N.I.test$p.value < 0.05)
%@
%
%Finally, the total number of nodes \gls{N} was consistently over-estimated by
%about a factor of two
%    (error d[,format(median(N_error), scientific=FALSE)] 
%    [d[,format(quantile(N_error, 0.25), scientific=FALSE)] - 
%     d[,format(quantile(N_error, 0.75), scientific=FALSE)]]).
%The fitted \gls{GLM} indicated that the true values of both \gls{alpha} and
%\gls{I} had an effect on the error in the estimated \gls{N}
%    ($p$-values $pp(N.glm[Parameter == "alpha", min(p)], eq=FALSE)$ and 
%     $pp(N.glm[Parameter == "I", min(p)], eq=FALSE)$,
%     \cref{tab:glmN}),
%but that the true value of \gls{m} did not. The error in the estimated \gls{N}
%when \gls{alpha} was equal to 1.5 was slightly lower than for other values of
%\gls{alpha}
%    (median [IQR] error rates 
%     d[true_alpha == 1.5, format(median(N_error), scientific=FALSE)] 
%    [d[true_alpha == 1.5, format(quantile(N_error, 0.25), scientific=FALSE)] - 
%     d[true_alpha == 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]]
%     for \gls{alpha} = 1.5 vs. 
%     d[true_alpha < 1.5, format(median(N_error), scientific=FALSE)] 
%    [d[true_alpha < 1.5, format(quantile(N_error, 0.25), scientific=FALSE)] - 
%     d[true_alpha < 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]]
%     for \gls{alpha} < 1.5),
%and this difference was statistically significant
%    (Wilcoxon rank-sum test, $p pp(N.alpha.test$p.value, eq=TRUE)$).
%Similarly, the error in the estimated \gls{N} when \gls{I} was equal to 3000 was
%lower than for other values of \gls{I}
%    (median [IQR] error rates 
%     d[true_I == 3000, format(median(N_error), scientific=FALSE)] 
%    [d[true_I == 3000, format(quantile(N_error, 0.25), scientific=FALSE)] - 
%     d[true_I == 3000, format(quantile(N_error, 0.75), scientific=FALSE)]]
%     for \gls{I} = 3000 vs. 
%     d[true_alpha < 3000, format(median(N_error), scientific=FALSE)] 
%    [d[true_alpha < 3000, format(quantile(N_error, 0.25), scientific=FALSE)] - 
%     d[true_alpha < 3000, format(quantile(N_error, 0.75), scientific=FALSE)]]
%     for \gls{I} < 3000),
%and this difference was statistically significant
%    (Wilcoxon rank-sum test, $p pp(N.I.test$p.value, eq=TRUE)$).
%
%\begin{figure}
%  \includegraphics{abc-point-estimate-m2}
%  \caption[\Acrlong{MAP} point estimates for \gls{BA} model parameters obtained
%    by running \software{netabc} on simulated data, for simulations with $m = 2$.] 
%  {
%    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by         
%    running \software{netabc} on simulated data. Values shown are for               
%    simulations with \gls{m} = 2. Dashed lines indicate true values. (A)            
%    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations  
%    against known values. (B) Estimates of \gls{m} and \gls{N} which were held   
%    fixed in these simulations at the values \gls{m} = 2 and \gls{N} = 5000. 
%  }
%  \label{fig:abcptm2}
%\end{figure}
%
%<<posterior_sims, echo=FALSE>>=
%    N <- 5000
%    replicate <- 0
%    lab <- NULL
%    for (m in c(2, 3, 4)) {
%    for (I in c(1000, 2000)) {
%    for (alpha in c(0, 0.5, 1, 1.5)) {
%        lab <- c(lab, sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate))
%    }
%    }
%    }
%    lab <- paste(lab, collapse=",")
%@
%
%\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
%distribution on the \gls{BA} parameters for one simulation. Equivalent plots
%for one replicate simulation with each combination of parameters can be found
%in \cref{lab}. \Gls{HPD} intervals around \gls{alpha} and \gls{I} were
%narrow relative to the region of nonzero prior density, whereas the intervals
%for $m$ and \gls{N} were widely dispersed. \Cref{tab:abchpd} shows point
%estimates and 95\% \gls{HPD} intervals averaged over all simulations.
%
%\begin{figure}
%  \includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
%  \caption[Approximate marginal posterior distributions of BA model parameters
%      obtained by applying \textit{netabc} to a simulated transmission tree
%      with values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]
%    {
%        Approximate marginal posterior distributions of BA model parameters
%        obtained by applying \textit{netabc} to a simulated transmission tree
%        with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ =
%        5000. Vertical dashed lines indicate true values. Shaded areas are 95\%
%        highest posterior density intervals. $x$-axes indicate regions of
%        nonzero prior density.
%    }
%  \label{fig:abcex}
%\end{figure}
%
%
%\begin{table}
%    \centering
%    \input{\tablepath/abc-hpd}
%    \caption[
%        Maximum \textit{a priori} estimates and 95\% highest posterior density
%        (HPD) interval boundaries for \gls{BA} model parameters estimated with
%        \software{netabc}, averaged over simulated transmission trees.
%    ]
%    {
%        Maximum \textit{a priori} estimates and 95\% highest posterior density
%        (HPD) interval boundaries for \gls{BA} model parameters estimated with
%        \software{netabc}, averaged over simulated transmission trees.
%    }
%    \label{tab:abchpd}
%\end{table}
%
%<<mixed_peerdriven, include=FALSE>>=
%  library(coda)
%  d <- fread("bzcat ../../simulations/abc-pa-mixed-alpha/abc/*")
%  d <- d[iter == max(iter)]
%  d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
%  params <- c("alpha", "m", "N", "I")
%  d <- melt(d, measure.vars=params, variable.name="parameter")
%  f <- function (x) {
%    dens <- density(x)
%    hpd <- HPDinterval(mcmc(x))
%    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
%         hpd.max=hpd[,"upper"])
%  }
%  d <- d[,f(value), by=parameter]
%  setkey(d, parameter)
%
%  m <- collect.metadata("../../simulations/abc-pa-peerdriven/abc/*")
%  p <- fread(paste("bzcat", rownames(m)[m$sample_peer == 1]))
%  p <- p[iter == max(iter)]
%  p <- p[sample(1:nrow(p), prob=weight, replace=TRUE)]
%  params <- c("alpha", "m", "N", "I")
%  p <- melt(p, measure.vars=params, variable.name="parameter")
%  f <- function (x) {
%    dens <- density(x)
%    hpd <- HPDinterval(mcmc(x))
%    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
%         hpd.max=hpd[,"upper"])
%  }
%  p <- p[,f(value), by=parameter]
%  setkey(p, parameter)
%  options(scipen=5)
%@
%
%To test the effect of model misspecification, we simulated one network where
%the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
%other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
%HPD] estimates for each parameter were: 
%$\alpha$, 
%  round(d["alpha", map], 2) 
%  [round(d["alpha", hpd.min], 2) -
%   round(d["alpha", hpd.max], 2)];
%$I$,
%  d["I", round(map)] 
%  [d["I", round(hpd.min)] -
%   d["I", round(hpd.max)]];
%$m$,
%  d["m", floor(map)] 
%  [d["m", floor(hpd.min)] -
%   d["m", floor(hpd.max)]];
%$N$,
%  d["N", round(map)] 
%  [d["N", round(hpd.min)] -
%   d["N", round(hpd.max)]].
%The approximate posterior distributions for this simulation are shown in
%\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
%transmission tree in a peer-driven fashion, where the probability to sample a
%node was twice as high if one of its peers had already been sampled. The
%parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
%$I$ = 2000. The estimated values were
%$\alpha$, 
%  p["alpha", round(map, 2)] 
%  [p["alpha", round(hpd.min, 2)] -
%   p["alpha", round(hpd.max, 2)]];
%$I$,
%  p["I", floor(map)] 
%  [p["I", floor(hpd.min)] -
%   p["I", floor(hpd.max)]];
%$m$,
%  p["m", floor(map)] 
%  [p["m", floor(hpd.min)] -
%   p["m", floor(hpd.max)]];
%$N$,
%  p["N", floor(map)] 
%  [p["N", floor(hpd.min)] -
%   p["N", floor(hpd.max)]].
%The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
%of these results were in line with estimates obtained on other simulated
%datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
%$\alpha$ was somewhat lower than typical.

%\subsubsection*{Effect of parameters on power-law exponent}
%
%\Cref{tab:glm} shows the estimated parameters for a log-link \gls{GLM} fitted
%to the observed distribution of \gls{gamma} values. The coefficients are
%interpretable as multiplicative effects.
%
%\begin{table}
%    \centering
%    \input{\tablepath/pa-gamma-glm}
%    \caption{Estimated \gls{GLM} parameters for relationship between power-law
%    exponent \gls{gamma} and \gls{BA} model parameters.}
%    \label{tab:glm}
%\end{table}

\section{Application to real world HIV data}
\label{sec:hiv}

\subsection{Methods}

{\color{blue}\uline{
The synthetic data experiments presented in the previous section demonstrated 
that the \gls{alpha} and \gls{I} parameters, and to a lesser extent the \gls{N}
parameter, had measurable impacts on tree shape and were identifiable with
kernel-assisted \gls{ABC}. Therefore, our next step was to use
\software{netabc} to estimate the parameters of the \gls{BA} model for real
world \gls{HIV} epidemics. }} Because the \gls{BA} model assumes a single
connected contact network, it is most appropriate to apply to groups of
individuals who are epidemiologically related. Therefore, we searched for
published \gls{HIV} datasets which originated from existing clusters, either
phylogenetically or geographically defined. In addition, we analysed an
in-house dataset sampled from \gls{HIV}-positive individuals in British
Columbia, Canada (the ``BC data''). The datasets are summarized in
\cref{tab:data}.

\begin{table}[ht]
  \centering
  \input{\tablepath/realdata-bc}
  \caption[Characteristics of published HIV datasets analyzed with \software{netabc}.]
  {
    Characteristics of published HIV datasets analyzed with \software{netabc}. 
    Abbreviations: MSM, men who have sex with men; HET, heterosexual; IDU,
    injection drug users. The \textcite{novitsky2013phylogenetic,novitsky2014impact} data
    were sampled from a primarily heterosexual risk environment, but did not
    explicitly exclude other risk factors.
  }
  \label{tab:data}
\end{table}

We downloaded all sequences associated with each published study from GenBank.
For the \textcite{novitsky2014impact} data, each \textit{env} sequence was
aligned pairwise to the HXB2 reference sequence (GenBank accession number
K03455) and the hypervariable regions were clipped out with
\software{BioPython} version 1.66+~\autocite{cock2009biopython}. Sequences were
multiply aligned using \software{MUSCLE} version 3.8.31
\autocite{edgar2004muscle}, and alignments were manually inspected with
\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
constructed from the nucleotide alignments by approximate maximum likelihood
using \software{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
\gls{GTR} model~\autocite{tavare1986some}. Transmission trees were estimated by
rooting and time-scaling the phylogenies by root-to-tip regression, using a
modified version of Path-O-Gen (distributed as part of
BEAST~\autocite{drummond2007beast}) as described
previously~\autocite{poon2015phylodynamic}.

Three of the datasets \autocite[][and the BC data]{li2015hiv,novitsky2014impact}
were initially much larger than the others, containing 1265, 1299, and 7923
sequences respectively. To ensure that the analyses were comparable, we reduced
these to a number of sequences similar to the smaller datasets. For the
\citeauthor{li2015hiv} and BC datasets, we detected clusters of size 280 and
399 respectively using a patristic distance cutoff of 0.02 as described
previously~\autocite{poon2015impact}. Only sequences within these clusters were
carried forward. For the \textcite{novitsky2014impact} data, no large clusters
were detected using the same cutoff, so we analysed a subtree of size 180
chosen arbitrarily.

Empirical studies of contact networks often report the exponent $\gamma$ of the
power law degree distribution. To compare our results to the literature, we
simulated 100 networks each according to the \gls{MAP} parameter estimates
obtained for each investigated dataset. The power-law exponent $\gamma$ was
calculated for each network using the \software{fit\_power\_law} function in
\software{igraph}, with the `R.mle' implementation. The median of the 100
$\gamma$ values was taken as a point estimate for the associated dataset.

For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2) and $N$ and
$I$ jointly uniform on the region \{$n \leq N \leq 10000$, $n \leq I \leq
10000$, $I \leq N$\}, where $n$ is the number of tips in the tree (see
\cref{tab:data}). Since the value $m = 1$ produces networks with no cycles,
which we considered fairly implausible, we ran one analysis with the prior $m
\sim$ DiscreteUniform(1, 5), and one with the prior $m \sim$ DiscreteUniform(2,
5). The other parameters to the SMC algorithm were the same as used for the
simulation experiments, except that we used 10000 particles instead of 1000 to
increase the accuracy of the estimated posterior. This was computationally
feasible due to the small number of runs required for this analysis.

We applied kernel-assisted ABC to five published HIV datasets (\cref{tab:data}),
and found substantial heterogeneity among the parameter estimates
(\cref{fig:abchpd,fig:abchpdm2}). Plots of the marginal posterior distributions
for each dataset are shown in
\cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
estimated $\alpha$ values near unity for the prior allowing $m = 1$ (\gls{MAP}
estimates [95\% \gls{HPD}] 
  0.76 
  [0.07 - 
   1.31]
and
  0.61 
  [0.01 -
   1.03] respectively).
The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
prior, although the credible interval of the \textcite{niculescu2015recent}
data was narrower
  (0.07 - 
   1.31).
When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
low estimated $\alpha$ values
  (0.31 
  [0 - 
  0.73]
and
  0.38 
  [0.02 -
   0.82]). 
However, the MAP estimates increased when $m = 1$ was not permitted, although
the HPD intervals remained roughly the same
  (0.55 
  [0.01 - 
  0.93]
and
  0.5 
  [0.07 -
   0.94]).
The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
for both priors on $m$
  (0.5 for $m \geq 1$;
   0.34 for $m \geq 2$).
However, the confidence interval was much wider when $m = 1$ was allowed
  ([0.03 -
    1.61] for $m \geq 1$ vs.
   [0 -
    0.73] for $m \geq 2$).

For all the datasets except \citeauthor{novitsky2014impact}, estimated values
of $I$ were below 2000 when $m = 1$ was allowed, with relatively narrow HPD
intervals compared to the nonzero prior density region
  (\citeauthor{cuevas2009hiv}, 943 
  [285 -
   2111];
   \citeauthor{niculescu2015recent}, 1032
  [136 - 
   2847];
  \citeauthor{li2015hiv}, 1437 
  [364 -
   2659];
   \citeauthor{wang2015targeting}, 872
  [173 - 
   2119]).
The \citeauthor{novitsky2014impact} data was the outlier, with a very high
estimated $I$, and HPD interval spanning almost the entire prior region
  (5586 
  [187 -
   8812]).
The $I$ estimates and HPD intervals were generally robust to the choice of
prior on $m$, with slightly narrower HPD intervals (compare
\cref{fig:abchpd,fig:abchpdm2}).

>>>>

{\color{blue}\uline{
To evaluate the effects of the true parameter values on the accuracy of the
posterior mean estimates, we analysed the \gls{alpha} and \gls{I} parameters
individually using \glspl{GLM}. The response variable was the error of the
point estimate, and the predictor variables were the true values of
\gls{alpha}, \gls{I}, and \gls{m}. We did not test for differences across true
values of \gls{N}, because \gls{N} was not varied in these simulations. The
distribution family and link function for the \glspl{GLM} were chosen as
Gaussian and inverse, respectively, by examination of residual plots and
\gls{AIC}. The $p$-values of the estimated \glspl{GLM} coefficients were
corrected using Holm-Bonferroni correction~\autocite{holm1979simple} with $n =
6$ (two \glspl{GLM} with three predictors each). Because there was clearly
little to no identifiability of the \gls{N} and \gls{m} parameters with
\gls{ABC} (see Results in next section), we did not construct \glspl{GLM} for
those parameters. }}

@@
